---
author: "Group 6"
date: "2025-11-24"
title: "Default of Credit Card Clients in Taiwan Dataset Analysis and Modeling using Logistic Regression in R"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The "Default of Credit Card Clients in Taiwan" dataset contains information about credit card clients in Taiwan, including their demographic details, credit history, and payment behavior. The primary objective of this analysis is to build a logistic regression model to predict whether a client will default on their credit card payment next month based on the available features.

# features definition

This research employed a binary variable, default payment (Yes = 1, No = 0), as the response variable. This study reviewed the literature and used the following 23 variables as explanatory variables: X1: Amount of the given credit (NT dollar): it includes both the individual consumer credit and his/her family (supplementary) credit. X2: Gender (1 = male; 2 = female). X3: Education (1 = graduate school; 2 = university; 3 = high school; 4 = others). X4: Marital status (1 = married; 2 = single; 3 = others). X5: Age (year). X6 - X11: History of past payment. We tracked the past monthly payment records (from April to September, 2005) as follows: X6 = the repayment status in September, 2005; X7 = the repayment status in August, 2005; . . .;X11 = the repayment status in April, 2005. The measurement scale for the repayment status is: -1 = pay duly; 1 = payment delay for one month; 2 = payment delay for two months; . . .; 8 = payment delay for eight months; 9 = payment delay for nine months and above. X12-X17: Amount of bill statement (NT dollar). X12 = amount of bill statement in September, 2005; X13 = amount of bill statement in August, 2005; . . .; X17 = amount of bill statement in April, 2005. X18-X23: Amount of previous payment (NT dollar). X18 = amount paid in September, 2005; X19 = amount paid in August, 2005; . . .;X23 = amount paid in April, 2005.

# Load Necessary Libraries

```{r load-libraries}
library(tidyverse)
library(readxl)
library(corrplot)
```

# Load the Dataset

```{r load-data}
df <- read_excel("datasets/default of credit card clients.xls", # nolint
                    skip = 1) # nolint

df |> glimpse()
df |> head()
```

Since the ID column is the row number for the dataset, hence it is not required for this analysis.

```{r remove-id-col}
df <- df |> select(-ID)

df |> head()
```

# Data Summary

```{r data-summary}
summary(df)
```

## Check for Missing Values

```{r check-missing}
colSums(is.na(df))
```

Looking at the missing values summary above, we can easily tell that there are no missing values in the dataset.

# Exploratory Data Analysis (EDA)

Distribution of the target variable

```{r eda}
df |> ggplot(aes(x = default.payment.next.month)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Default Payment Next Month",
       x = "Default Payment Next Month",
       y = "Count")
```

Correlation between numerical features
```{r correlation-plot}
library(corrplot)